Fit by design: Developing seed-substrate combinations to adapt dike grasslands to microclimate

### 2023-07-04 manuscript.Rmd
Update manuscript.Rmd

1.  Sowing is a well-established restoration technique to overcome dispersal limitation. Seed mixtures adapted to certain environmental conditions like substrate or microclimate are most effective to achieve functional communities. This is especially important if the restored vegetation has to protect critical infrastructure like roadsides and dikes. Here, an improved seed--substrate combination will secure slope stability, make restorations more effective, and generate species-rich grasslands.
2.  A factorial field experiment addressed this topic on a dike at River Danube in SE Germany in 2018--2021. Within 288 plots, we tested three sand admixtures, two substrate depths, two seed densities and two seed mixture types (mesic hay meadow, semi-dry calcareous grassland) in north and south exposition, and measured the recovery completeness by calculating the successional distance to reference sites, the persistence of sown species, and the Favourable Conservation Status (FCS) of target species.
3.  Overall, the sown vegetation developed in the desired direction, but a recovery debt remained after four years, and some plots still showed similarities to negative references from ruderal sites. In north exposition, hay meadow-seed mixtures developed closer to the respective reference communities than dry-grassland mixtures.
4.  In south exposition, the sown communities developed poorly which might be due to a severe drought during establishment. This initial negative effect remained over the entire observation period.
5.  Sand admixture had a slightly positive effect on target variables, while the tested substrate depths, seed densities and mixture types had no effects on species persistence or FCS.
6.  Synthesis and applications: Site-adapted seed mixtures make restoration more effective. However, applying several seed--substrate combinations might foster beta diversity. Furthermore, additional management efforts are recommended, as they might be necessary to reduce the recovery debt, as well as re-sowing after unfavourable conditions. \clearpage \newpage

@@

Dikes are promising sites for the restoration of species-rich grasslands because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in agricultural landscapes [@Batori.2020]. In addition, dikes are interesting for experiments with seed-substrate combinations because the combinations can be tested on contrasting microclimates of the steep slopes (\>1:3) with different expositions.

contrasting substrate layers and dense swards for erosion protection characterise these habitats [@Batori.2016; @berendse2015; @husicka2003]. Dikes can reconcile several ecosystem functions including both flood security and rich biodiversity [@teixeira2022], which can be fostered by an adapted seed--substrate combination.

@@

2.  Nutrient reduction by sand addition and shallow substrates increase the likeliness of the establishment of dry-grassland seed mixtures compared to mesic ones.

3.  High seed densities and reduced soil fertility increase the likeliness of the establishment of sown plants and suppress non-target species compared to low density sowing.

@@

The vegetation in south exposition had a more ruderal and xerophytic species compositions than in north exposition, which contrasts @Batori.2016 who found different compositions in south exposition only for riverside slopes which caused a more mesotrophic vegetation. The differences in our experiment might be due to methodical reasons, since the geotextile, which had been implemented on the southern slope, was removed after two weeks. This was unfortunate for at least some seedlings, and amplified by the intense drought in summer 2018 and 2019 [cf. @hari2020; @Larson.2021; @orrock2023]. The lasting negative effect on persistence and FCS on the southern slope suggests a legacy effect of adverse weather conditions after sowing as observed by other studies [@Groves.2020; @atkinson2023]. These conditions during the establishment phase might have led to a special trajectory [@suding2004], and probably leveled the distinction of the seed mixture types in south exposition.

@@

Our aim was to identify perfect seed--substrate combinations regarding restoration effectiveness and biodiversity (H2, H3). For evaluating effectiveness, we measured the persistence of the sown species, and FCS for investigating plant biodiversity. However, we could not identify an interaction effect for neither one of these indices. We would have expected a better performance of hay-meadow seed mixtures with lower sand admixture and for dry grasslands with higher sand admixture (H2). Our results suggest that, at least after four years, the substrate conditions are within the range of both seed mixture types (hay meadows vs dry grasslands). Although, both types are clearly phytosociologically and functionally distinct, they are still relatively close, because they contain shared species and develop under similar site conditions with modified sub-associations [Appendix A3, @husicka2003; @Oberdorfer.1993]. Other grassland studies could identify more or less clear interactions of opposing habitat preferences or functional traits along the gradients of productivity, moisture and nutrients [@Zirbel.2020; @Freitag.2021; @Kaulfuss.2022]. However, these studies did not work with an experimental set up of different seed--substrate combinations, but analysed the result of habitat and biotic filtering after 1, 5 and 15yrs, respectively. Furthermore, the non-existence of ideal combinations could be explained by priority effects that means that the species of the imperfect-adapted seed mixture type could establish earlier and pre-empted the available niches for the species of related habitat types [@Fukami.2015].

>>>>

1.  Sowing is a well-established restoration technique to overcome dispersal limitation. Seed mixtures adapted to certain environmental conditions like substrate or microclimate are most effective to achieve functional communities. This is especially important if the restored vegetation has to protect critical infrastructure like roadsides and dikes. Here, an improved seed--substrate combination will secure slope stability, make restorations more effective, and generate species-rich grasslands.
2.  A factorial field experiment addressed this topic on a dike at River Danube in SE Germany in 2018--2021. Within 288 plots, we tested three sand admixtures, two substrate depths, two seed densities and two seed mixture types (mesic hay meadow, semi-dry calcareous grassland) in north and south exposition, and measured the recovery completeness by calculating the successional distance to reference sites, the persistence of sown species, and the Favourable Conservation Status (FCS) of target species.
3.  Overall, the sown vegetation developed in the desired direction, but a recovery debt remained after four years, and some plots still showed similarities to negative references from ruderal sites. In north exposition, hay meadow-seed mixtures developed closer to the respective reference communities than dry-grassland mixtures.
4.  In south exposition, the sown communities developed poorly which might be due to a severe drought during establishment. This initial negative effect remained over the entire observation period.
5.  Sand admixture had a slightly positive effect on target variables, while the tested substrate depths, seed densities and mixture types had no effects on species persistence or FCS.
6.  Synthesis and applications: Site-adapted seed mixtures make restoration more effective. However, applying several seed--substrate combinations might foster beta diversity. Furthermore, additional management efforts are recommended, as they might be necessary to reduce the recovery debt, as well as re-sowing after unfavourable conditions.

@@

Dikes are promising sites for the restoration of species-rich grasslands because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in agricultural landscapes [@Batori.2020]. Several ecosystem functions can be reconciled by dike grasslands like erosion control [@husicka2003; @berendse2015] and biodiversity [@teixeira2022], which can be fostered by an adapted seed--substrate combination. Experiments with such seed-substrate combinations on dikes benefit from contrasting microclimates of the different expositions of the steep slopes (\>1:3) [@suggitt2010] .

@@

2.  Nutrient reduction by sand addition and shallow substrates improve the establishment of dry-grassland seed mixtures compared to mesic ones.

3.  High seed densities and reduced soil fertility improve the establishment of sown plants and suppress non-target species compared to low density sowing and high soil fertility.

@@

The vegetation in south exposition had a more ruderal and xerophytic species compositions than in north exposition, which contrasts @Batori.2016 who found different compositions in south exposition only for riverside slopes which caused a more mesotrophic vegetation. The differences in our experiment might be due to methodical reasons, since the geotextile, which had been implemented on the southern slope, was removed after two weeks. This was unfortunate for at least some seedlings, and amplified by the intense drought in summer 2018 and 2019 [cf. @hari2020; @Larson.2021; @orrock2023]. The lasting negative effect on persistence and FCS on the southern slope suggests a legacy effect of adverse weather conditions after sowing as observed by other studies [@Groves.2020; @atkinson2023]. These conditions during the establishment phase might have led to a special trajectory [@suding2004], and probably levelled the distinction of the seed mixture types in south exposition.

@@

Our aim was to identify perfect seed--substrate combinations regarding restoration effectiveness and biodiversity (H2, H3). For evaluating effectiveness, we measured the persistence of the sown species, and FCS for investigating plant biodiversity. However, we could not identify an interaction effect for neither one of these indices. We would have expected a better performance of hay-meadow seed mixtures with lower sand admixture and for dry grasslands with higher sand admixture (H2). Our results suggest that, at least after four years, the substrate conditions are within the range of both seed mixture types (hay meadows vs dry grasslands). Although, both types are clearly phytosociologically and functionally distinct, they are still relatively close, because they contain shared species and develop under similar site conditions with modified sub-associations [Appendix A3, @husicka2003; @Oberdorfer.1993]. Other grassland studies could identify more or less clear interactions of opposing habitat preferences or functional traits along the gradients of productivity, moisture and nutrients [@Zirbel.2020; @Freitag.2021; @Kaulfuss.2022]. However, these studies did not work with an experimental set up of different seed--substrate combinations, but analysed the result of habitat and biotic filtering after 1, 5 and 15yrs, respectively. Furthermore, the non-existence of ideal combinations could be explained by priority effects that means that the species of the imperfect-adapted seed mixture type could establish earlier and pre-empted the available niches for the species of related habitat types [@Fukami.2015]. \clearpage \newpage
